#DNA Methylation Analysis R - analysis with integrated beta-values

```{r message=FALSE, warning=FALSE, results = 'hide'}
version
if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install()
```

Load the libraries

```{r message=FALSE, warning=FALSE, results = 'hide'}
#loading those libraries
library(gt)
library(tidyverse)
library(RColorBrewer)
library(dplyr)
library(readr)
library(ggrepel)
library(limma)
library(GenomicRanges)
library(envalysis)
library(scatterplot3d)
library(vegan)
select <- dplyr::select

```

##Read merged beta values and reference file and do PCAs

The below code shows how SHK482 is an outlier and should be discarded
If you want to work with the data without the outlier, run the code chunk after this
For the Y chromosome removed data and annotation file go to that code chunk 
```{r}
setwd("/Users/egeezen/Desktop/PhD/2025 First Term/DNA Methylation/")
baseDir <- "/Users/egeezen/Desktop/PhD/2025 First Term/DNA Methylation/"

#read the patients metadata 
patients<- read_csv("patients_new.csv", col_types=cols(.default = "c")) 

#read the RDS object containing beta values and annotation file
rds_list_final <- readRDS("merged_beta_values_and_annotation_file.rds")
merged_beta_values <- rds_list_final$merged_beta_values
annotation_file <- rds_list_final$annotation_file

#do pca analysis and get as a dataframe ->using plotMDS function from limma
construct_pca_dataframe<- function (merged_beta_values, patients){
  #the parameters are optimized for our microarray results
  mds_res <- plotMDS(merged_beta_values,top = 10000, gene.selection = "pairwise", plot = FALSE)
  leading_var <- mds_res$var.explained  # These are already percentages
  percentages <- leading_var[1:2]
  percentages <- round(100 *percentages)
  # Extract coordinates
  mds_df <- data.frame(
    Sample = colnames(merged_beta_values),
    Dim1 = mds_res$x,
    Dim2 = mds_res$y
  )
  mds_df <- mds_df %>%
    rename("Sample" = "Patient_Name") %>%
    left_join( patients,by = "Patient_Name")
  return(list(mds_df = mds_df, variance = percentages))
}

pca_results <- construct_pca_dataframe(merged_beta_values, patients)

#construct the PCoA (Principle Coordinate Analysis) plot
group_colors <- c("Healthy" = "#C1CFA1",
                  "VeRA" = "#F7CFD8",
                  "RA" = "#A888B5",
                  "OA" = "#C5D3E8")

sex_colors <- c("M" = "#E9F5BE",
                "F" = "#F1BA88")

pca_results$mds_df[,"Final_Group"] = factor(pca_results$mds_df[,"Final_Group"] , levels = c("Healthy","VeRA","RA","OA"))

ggplot(pca_results$mds_df, aes(x = Dim1, y = Dim2, fill = Final_Group)) +
  geom_point(size = 3, alpha = 1 ,shape = 21, stroke = 0.3, color = "black" ) +
  scale_fill_manual(values = group_colors) + 
  xlab(paste0("PC1 (", pca_results$variance[1], "%)")) +
  ylab(paste0("PC2 (", pca_results$variance[2], "%)")) +
  theme_publish() +
  coord_fixed() +
  labs(fill = "")


ggsave(file = "all_before_x_y_removal.png",
       plot = last_plot(),
       dpi = 300, 
       bg = "transparent", 
       width = 5, height =5)

```

##Without the outlier from now on!, exploratory analysis using PCAs

```{r}
setwd("/Users/egeezen/Desktop/PhD/2025 First Term/DNA Methylation/")
baseDir <- "/Users/egeezen/Desktop/PhD/2025 First Term/DNA Methylation/"

#read the patients metadata 
patients<- read_csv("patients_new.csv", col_types=cols(.default = "c")) 
patients<- patients[-43,] #get rid of the outlier 

#scale the age for stability
patients$age_z <- scale(as.numeric(patients$age))

#read the RDS object containing beta values and annotation file
rds_list_final <- readRDS("merged_beta_values_and_annotation_file.rds")
merged_beta_values <- rds_list_final$merged_beta_values
annotation_file <- rds_list_final$annotation_file
merged_beta_values$SHK482 <- NULL #get rid of the outlier 

#put the location information
patients <- patients %>%
  mutate(final_location = case_when(
    location == "thumb" ~ "hand",
    location == "PIP"   ~ "hand",
    location == "wrist"   ~ "hand",
    location == "finger"   ~ "hand",
    TRUE ~ location))

#do pca analysis and get as a dataframe ->using plotMDS function from limma
construct_mds_dataframe <- function(merged_beta_values, patients, ndim = 6) {
  
  mds_res <- plotMDS(
    merged_beta_values,
    top = 10000,
    gene.selection = "pairwise",
    ndim = ndim,
    plot = FALSE
  )
  #calculate the coordinates
  vectors <- mds_res$eigen.vectors[, 1:ndim]
  values  <- mds_res$eigen.values[1:ndim]
  mds_coords <- t(t(vectors) * sqrt(values)) #eigen calculation
  colnames(mds_coords) <- paste0("PC", 1:ndim)
  
  # variance explained (already proportions)
  var_explained <- round(100 * mds_res$var.explained[1:ndim], 1)
  
  mds_df <- data.frame(
      Patient_Name = colnames(merged_beta_values), # Ensures row/col alignment
      mds_coords)
  
  mds_df <- mds_df %>%
    left_join(patients, by = "Patient_Name")
  
  return(list(
    mds_df = mds_df,
    variance = var_explained
  ))
}

mds_results <- construct_mds_dataframe(merged_beta_values, patients, ndim = 6)

#construct the PCoA (Principle Coordinate Analysis) plot
group_colors <- c("Healthy" = "#C1CFA1",
                  "VeRA" = "#F7CFD8",
                  "RA" = "#A888B5",
                  "OA" = "#C5D3E8")

sex_colors <- c("M" = "#E9F5BE",
                "F" = "#F1BA88")

location_colors <- c(
  "shoulder" = "#D6CDEA", 
  "hand"     = "#FDD9C3", 
  "knee"     = "#B5EAD7",  
  "ankle"    = "#E9F5BE" )

mds_results$mds_df$final_location <- factor(patients$final_location, 
                                        levels = c("shoulder", "hand", "knee", "ankle"))

mds_results$mds_df[,"Final_Group"] = factor(mds_results$mds_df[,"Final_Group"] , levels = c("Healthy","VeRA","RA","OA"))

#to plot the pca plot pairs
plot_mds_pair <- function(df, pc_x, pc_y, variance, fill_var, fill_colors) {
  
  ggplot(df, aes_string(x = pc_x, y = pc_y, fill = fill_var)) +
    geom_point(
      size = 3,
      shape = 21,
      color = "black",
      stroke = 0.3
    ) +
    scale_fill_manual(values = fill_colors) +
    xlab(paste0(pc_x, " (", variance[as.numeric(sub("PC","",pc_x))], "%)")) +
    ylab(paste0(pc_y, " (", variance[as.numeric(sub("PC","",pc_y))], "%)")) +
    coord_fixed() +
    theme_publish() +
    labs(fill = "")
}

#PC1 & PC2 - Joint 
p1 <- plot_mds_pair(
  mds_results$mds_df,
  "PC1", "PC2",
  mds_results$variance,
  "final_location",
  location_colors
)

#save the plot
ggsave(file = "location_all_pcoa_p1_p2.png", plot = p1, dpi = 300, bg = "transparent", 
        width = 5, height =5)

ggsave(file = "location_all_pcoa_p1_p2.pdf", plot = p1, dpi = 300,
        width = 5, height =5)

#PC2 and PC3 - Sexes
p3 <- plot_mds_pair(
  mds_results$mds_df,
  "PC3", "PC4",
  mds_results$variance,
  "sex",
  sex_colors)

ggsave(file = "sexes_all_pcoa_p3_p4.png", plot = p3, dpi = 300, bg = "transparent", 
        width = 5, height =5)

ggsave(file = "sexes_all_pcoa_p3_p4.pdf", plot = p3, dpi = 300,
        width = 5, height =5)

```

### for 3D-PCA plot

```{r}
#3D plot

# 1. Open the PDF device
pdf("3d_plot_p1p2p3_disease.pdf", width = 6, height = 6)

# 2. Match the the colors
fill_colors <- group_colors[as.character(mds_results$mds_df$Final_Group)]

# 3. Create the plot
s3d <- scatterplot3d(
  x = mds_results$mds_df$PC3,
  y = mds_results$mds_df$PC1,
  z = mds_results$mds_df$PC2,
  pch = 21,
  color = "black",
  bg = fill_colors,
  lwd = 0.5,
  cex.symbols = 1.5,     
  ylab = "PC1 (12.5%)",
  zlab = "PC2 (8%)",
  xlab = "PC3 (5.2%)",
  angle = 45)

# 4. Add a legend
legend("topright", 
       legend = names(group_colors), 
       col = group_colors, 
       pch = 19, 
       title = "Disease",
       bty = "n") # No box around legend

# 5. Close the device to save the file
dev.off()

```

###Linear regression on PCs 

```{r}

pc_lm_results <- list(
  PC1 = summary(lm(PC1 ~ final_location + Final_Group + sex + age_z,
                   data = mds_results$mds_df)),
  PC2 = summary(lm(PC2 ~ final_location + Final_Group + sex + age_z , 
                   data = mds_results$mds_df)),
  PC3 = summary(lm(PC3 ~ final_location + Final_Group + sex + age_z, 
                   data = mds_results$mds_df)),
  PC4 = summary(lm(PC4 ~ final_location + Final_Group + sex + age_z , 
                   data = mds_results$mds_df))
)

pc_lm_results$PC1
pc_lm_results$PC2
pc_lm_results$PC3
pc_lm_results$PC4

```

###PERMANOVA

```{r}

# calculate distance matrix
dist_mat <- dist(t(merged_beta_values), method = "euclidean")

# 1. Run the marginal model
# This calculates the unique contribution of each variable

perm_result <- adonis2(
  dist_mat ~ final_location + Final_Group + sex + age_z,
  data = patients,
  permutations = 999,
  by = "margin"
)

# 2. Create a clean "Variance Explained" table
var_table <- data.frame(
  Variable = rownames(perm_result),
  Variance_Percent = round(perm_result$R2 * 100, 2), # Convert to %
  P_Value = perm_result$`Pr(>F)`
)

# Remove the 'Total' row to avoid confusion
var_table <- var_table[rownames(var_table) != "Total", ]

# Sort by highest contribution
var_table <- var_table[order(-var_table$Variance_Percent), ]

# Print the result
print(var_table)

# Prepare plotting table (exclude Residual and Total)
plot_table <- var_table %>%
  filter(!Variable %in% c("Residual", "Total")) %>%
  mutate(
  Variable = reorder(Variable, Variance_Percent),    
  Significance = ifelse(P_Value < 0.05, "Significant", "Not significant"),
    alpha_val = ifelse(P_Value < 0.05, 0.9, 0.35),
    p_text = case_when(
      P_Value <= 0.001 ~ "p < 0.001",
      P_Value < 0.05  ~ paste0("p = ", sprintf("%.3f", P_Value)),
      TRUE            ~ "ns"),
    label_text = paste0(sprintf("%.1f", Variance_Percent), "% (", p_text, ")")
  )

# Adjust the names on the right to match exactly what you want to appear
clean_labels <- c(
  "Final_Group" = "Diagnosis",
  "final_location" = "Joint", 
  "sex" = "Sex",
  "age_z" = "Age"
)

# Plot
p<- ggplot(plot_table, aes(x = Variable, y = Variance_Percent,
                           fill = Variable, alpha = alpha_val)) +

  geom_col(color = "black", width = 0.8, show.legend = FALSE) +
  
  geom_text(
    aes(label = label_text), 
    hjust = -0.1,
    size = 3.5,
    color = "black",
    fontface = "bold"
  ) +
  
  scale_alpha_identity() +
  
  scale_y_continuous(expand = expansion(mult = c(0, 0.5))) +
  
  scale_x_discrete(labels = clean_labels) +
  
  scale_fill_brewer(palette = "Pastel2") + 
  coord_flip() +
  
  labs(
    y = "Independent Variance Explained (%)", 
    x = NULL, # No need for x-label if categories are self-explanatory
    title = "Contribution of Metadata to Variance"
  ) +
  
  theme_publish(base_size = 14, base_linewidth = 0.7)  

ggsave(file = "permanova_marginal.pdf", plot = p, dpi = 300, 
        width = 5, height =4)

```

###PERMANOVA without X and Y chromosomes

```{r}
#we can pick the X and Y chromosomes from the annotation file
keep <- !(annotation_file$chr %in% c("chrX","chrY"))
annotation_file_filtered <- annotation_file[keep,]
#now we can filter our m_values by removing sex chromosome loci, this will reduce variability significantly
merged_beta_values_filtered  <- merged_beta_values[keep,]

# calculate distance matrix
dist_mat <- dist(t(merged_beta_values_filtered), method = "euclidean")

# 1. Run the marginal model
# This calculates the unique contribution of each variable

perm_result <- adonis2(
  dist_mat ~ final_location + Final_Group + sex + age_z,
  data = patients,
  permutations = 999,
  by = "margin"
)

# 2. Create a clean "Variance Explained" table
var_table <- data.frame(
  Variable = rownames(perm_result),
  Variance_Percent = round(perm_result$R2 * 100, 2), # Convert to %
  P_Value = perm_result$`Pr(>F)`
)

# Remove the 'Total' row to avoid confusion
var_table <- var_table[rownames(var_table) != "Total", ]

# Sort by highest contribution
var_table <- var_table[order(-var_table$Variance_Percent), ]

# Print the result
print(var_table)

# Prepare plotting table (exclude Residual and Total)
plot_table <- var_table %>%
  filter(!Variable %in% c("Residual", "Total")) %>%
  mutate(
  Variable = reorder(Variable, Variance_Percent),    
  Significance = ifelse(P_Value < 0.05, "Significant", "Not significant"),
    alpha_val = ifelse(P_Value < 0.05, 0.9, 0.35),
    p_text = case_when(
      P_Value <= 0.001 ~ "p < 0.001",
      P_Value < 0.05  ~ paste0("p = ", sprintf("%.3f", P_Value)),
      TRUE            ~ "ns"),
    label_text = paste0(sprintf("%.1f", Variance_Percent), "% (", p_text, ")")
  )

# Adjust the names on the right to match exactly what you want to appear
clean_labels <- c(
  "Final_Group" = "Diagnosis",
  "final_location" = "Joint", 
  "sex" = "Sex",
  "age_z" = "Age"
)

# Plot
p<- ggplot(plot_table, aes(x = Variable, y = Variance_Percent,
                           fill = Variable, alpha = alpha_val)) +

  geom_col(color = "black", width = 0.8, show.legend = FALSE) +
  
  geom_text(
    aes(label = label_text), 
    hjust = -0.1,
    size = 3.5,
    color = "black",
    fontface = "bold"
  ) +
  
  scale_alpha_identity() +
  
  scale_y_continuous(expand = expansion(mult = c(0, 0.5))) +
  
  scale_x_discrete(labels = clean_labels) +
  
  scale_fill_brewer(palette = "Pastel2") + 
  coord_flip() +
  
  labs(
    y = "Independent Variance Explained (%)", 
    x = NULL, # No need for x-label if categories are self-explanatory
    title = "Contribution of Metadata to Variance"
  ) +
  
  theme_publish(base_size = 14, base_linewidth = 0.7)  

ggsave(file = "permanova_marginal_without_x_y.pdf", plot = p, dpi = 300, 
        width = 5, height =4)


```

##Removing X and Y chromosomes and explore again

```{r}
#we can pick the X and Y chromosomes from the annotation file
keep <- !(annotation_file$chr %in% c("chrX","chrY"))
annotation_file_filtered <- annotation_file[keep,]
#now we can filter our m_values by removing sex chromosome loci, this will reduce variability significantly
merged_beta_values_filtered  <- merged_beta_values[keep,]

# you can do subsetting
# patients_subset <- patients %>% filter(final_location =="hand")
# subset_beta_values <- merged_beta_values_filtered[,patients_subset$Patient_Name]

#get pca
pca_results_filtered <- construct_pca_dataframe(merged_beta_values_filtered, patients)

#construct the PCoA (Principle Coordinate Analysis) plot
pca_results_filtered$mds_df$final_location <- factor(patients$final_location, 
                                            levels = c("shoulder", "hand", "knee", "ankle"))

pca_results_filtered$mds_df[,"Final_Group"] = factor(pca_results_filtered$mds_df[,"Final_Group"] , levels = c("Healthy","VeRA","RA","OA"))

#in this plot you can change the fill and scale_fill_manual labels to look
p<-ggplot(pca_results_filtered$mds_df, aes(x = Dim1, y = Dim2, fill = final_location)) + #change the fill argument here   
  geom_point(size = 3, alpha = 1 ,shape = 21, stroke = 0.3, color = "black" ) +
  scale_fill_manual(values = location_colors) + #change the values here
  xlab(paste0("PC1 (", pca_results_filtered$variance[1], "%)")) +
  ylab(paste0("PC2 (", pca_results_filtered$variance[2], "%)")) +
  #geom_label_repel(aes(label = sex), size = 3,fill=NA) +
  theme_publish() +
  coord_fixed() +
  labs(fill = "")

#save the plot
ggsave(file = "after_x_y_removal_location_all_pcoa.png",
       plot = p,
       dpi = 300, 
       bg = "transparent", 
       width = 5, height =5)

ggsave(file = "after_x_y_removal_location_all_pcoa.pdf",
       plot = p,
       dpi = 300, 
       width = 5, height =5)

# #before saving the annotation file, getting the unique names for genes
annotation_file_filtered$genesUniq <- sapply(annotation_file_filtered$UCSC_RefGene_Name, function(x) {
  paste(unique(strsplit(x, ";")[[1]]), collapse = ";")
})
#the final beta values and annotation files are saved
after_xy_removal_list <- list(merged_beta_values = merged_beta_values_filtered, annotation_file = annotation_file_filtered, patients = patients)

saveRDS(after_xy_removal_list, "merged_beta_values_and_annotation_file_after_filtered_xy.rds")

```

##Look at the regions that are open across the Patients (RA Knee, OA Knee, Healthy Knee, RA Hand, OA Thumb, OA Shoulder, RA Shoulder) and within joint locations

```{r}
#-----------------------------------------------------------
# Load ChromHMM tracks
#-----------------------------------------------------------
ra_knee_chromhmm_df      <- read.table("chromhmm tracks/ra_knee_368.bed", sep="\t", header=FALSE, stringsAsFactors=FALSE)
oa_knee_chromhmm_df      <- read.table("chromhmm tracks/oa_knee_335.bed", sep="\t", header=FALSE, stringsAsFactors=FALSE)
healthy_knee_chromhmm_df <- read.table("chromhmm tracks/healthy_knee_88.bed", sep="\t", header=FALSE, stringsAsFactors=FALSE)
ra_hand_chromhmm_df      <- read.table("chromhmm tracks/female_ra_hand_316.bed", sep="\t", header=FALSE, stringsAsFactors=FALSE)
oa_thumb_chromhmm_df     <- read.table("chromhmm tracks/oa_thumb_343.bed", sep="\t", header=FALSE, stringsAsFactors=FALSE)
ra_shoulder_chromhmm_df  <- read.table("chromhmm tracks/ra_shoulder_341.bed", sep="\t", header=FALSE, stringsAsFactors=FALSE)
oa_shoulder_chromhmm_df  <- read.table("chromhmm tracks/shoulder_oa_384.bed", sep="\t", header=FALSE, stringsAsFactors=FALSE)

#-----------------------------------------------------------
# ChromHMM state mapping
#-----------------------------------------------------------
chromhmm_state_map <- data.frame(
  state = paste0("E", 1:18),
  label = c(
    "Active TSS","Flanking Active TSS","Transcription at gene 5′ and 3′",
    "Strong transcription","Weak transcription","Genic enhancer",
    "Enhancer 1","Enhancer 2","Bivalent/Poised TSS","Bivalent enhancer",
    "Repressed PolyComb","ZNF genes & repeats","Heterochromatin",
    "Quiescent/Low","Repetitive/Quiescent","Weak promoter/poised",
    "Weak enhancer","Repressed/CTCF"
  ),
  stringsAsFactors = FALSE
)

open_states <- c(
  "Active TSS","Flanking Active TSS","Genic enhancer","Enhancer 1","Enhancer 2",
  "Strong transcription","Weak transcription","Weak enhancer","Weak promoter/poised",
  "Transcription at gene 5′ and 3′"
)

#-----------------------------------------------------------
# Helper: make GRanges for open chromatin
#-----------------------------------------------------------
make_open_gr <- function(chrom_df) {
  colnames(chrom_df) <- c("chr","start","end","state")
  chrom_df2 <- chrom_df %>% left_join(chromhmm_state_map, by="state")
  open_df <- chrom_df2 %>% filter(label %in% open_states)
  GRanges(seqnames=open_df$chr, ranges=IRanges(open_df$start, open_df$end))
}

# Create GRanges for all datasets
open_ra_knee      <- make_open_gr(ra_knee_chromhmm_df)
open_oa_knee      <- make_open_gr(oa_knee_chromhmm_df)
open_healthy_knee <- make_open_gr(healthy_knee_chromhmm_df)
open_ra_hand      <- make_open_gr(ra_hand_chromhmm_df)
open_oa_thumb     <- make_open_gr(oa_thumb_chromhmm_df)
open_ra_shoulder  <- make_open_gr(ra_shoulder_chromhmm_df)
open_oa_shoulder  <- make_open_gr(oa_shoulder_chromhmm_df)

#-----------------------------------------------------------
# Combine site-specific and all-sample open chromatin regions
#-----------------------------------------------------------

suppressWarnings(open_union_all <- reduce(c(open_ra_knee, open_oa_knee, open_healthy_knee, open_ra_hand, open_oa_thumb, open_ra_shoulder, open_oa_shoulder)) )
suppressWarnings(open_union_knee   <- reduce(c(open_ra_knee, open_oa_knee, open_healthy_knee)))
suppressWarnings(open_union_hand     <- reduce(c(open_ra_hand, open_oa_thumb)))
suppressWarnings(open_union_shoulder <- reduce(c(open_ra_shoulder, open_oa_shoulder)))

#-----------------------------------------------------------
# Probe annotation and GRanges
#-----------------------------------------------------------
probe_df <- annotation_file_filtered %>%
  rownames_to_column("probe") %>%
  select(probe, chr, pos) %>%
  filter(!is.na(chr) & !is.na(pos))

probe_gr <- GRanges(seqnames=probe_df$chr,
                    ranges=IRanges(probe_df$pos, width=1),
                    probe=probe_df$probe)

#-----------------------------------------------------------
# Helper: get overlapping probes per site
#-----------------------------------------------------------
get_open_probes <- function(open_union, probe_gr) {
  hits <- findOverlaps(probe_gr, open_union)
  unique(mcols(probe_gr)$probe[queryHits(hits)])
}

probes_all      <- get_open_probes(open_union_all, probe_gr)
probes_knee     <- get_open_probes(open_union_knee, probe_gr)
probes_hand     <- get_open_probes(open_union_hand, probe_gr)
probes_shoulder <- get_open_probes(open_union_shoulder, probe_gr)

# Save probe sets
saveRDS(probes_all,      "chromhmm tracks/open_probes_all_sites_xy_removed.rds")
saveRDS(probes_knee,     "chromhmm tracks/open_probes_knee_xy_removed.rds")
saveRDS(probes_hand,     "chromhmm tracks/open_probes_hand_xy_removed.rds")
saveRDS(probes_shoulder, "chromhmm tracks/open_probes_shoulder_xy_removed.rds")

#-----------------------------------------------------------
# Helper: extract site-specific beta and annotation
#-----------------------------------------------------------
get_site_data <- function(probes_site) {
  beta_site <- merged_beta_values_filtered[rownames(merged_beta_values_filtered) %in% probes_site, , drop=FALSE]
  annot_site <- annotation_file_filtered[rownames(annotation_file_filtered) %in% probes_site, , drop=FALSE]
  list(beta=beta_site, annot=annot_site)
}

site_data_all      <- get_site_data(probes_all)
site_data_knee     <- get_site_data(probes_knee)
site_data_hand     <- get_site_data(probes_hand)
site_data_shoulder <- get_site_data(probes_shoulder)

#-----------------------------------------------------------
# PCA function with patient filtering
#-----------------------------------------------------------
run_site_pca_plot <- function(site_name, site_data, patients, colors, site_label=NULL) {
  # Filter patient metadata if a specific site is provided
  if (!is.null(site_label)) {
    patients_sub <- patients %>% filter(final_location == site_label)
    beta_values_filt <- site_data$beta[,colnames(site_data$beta) %in% patients_sub$Patient_Name]
        # Run PCA
    pca <- construct_pca_dataframe(beta_values_filt, patients_sub)
    pca$mds_df$final_location <- factor(patients_sub$final_location,
                                      levels=c("shoulder", "hand", "knee", "ankle"))
    pca$mds_df$Final_Group <- factor(patients_sub$Final_Group,
                                       levels=c("Healthy","VeRA","RA","OA"))

    # Build plot
    plt <- ggplot(pca$mds_df, aes(x=Dim1, y=Dim2, fill=Final_Group)) +
      geom_point(size=3, alpha=1, shape=21, stroke=0.3, color="black") +
      scale_fill_manual(values=colors) +
      xlab(paste0("PC1 (", pca$variance[1], "%)")) +
      ylab(paste0("PC2 (", pca$variance[2], "%)")) +
      theme_publish() +
      coord_fixed() +
      labs(fill="")
  
    # Save
    ggsave(paste0("chromhmm tracks/pca_", site_name, "_open_regions.png"),
           plot=plt, dpi=300, width=5, height=5, bg="transparent")
    ggsave(paste0("chromhmm tracks/pca_", site_name, "_open_regions.pdf"),
           plot=plt, dpi=300, width=5, height=5)

  } else {
    patients_sub <- patients
    # Run PCA
    pca <- construct_pca_dataframe(site_data$beta, patients_sub)
    pca$mds_df$final_location <- factor(patients_sub$final_location,
                                      levels=c("shoulder", "hand", "knee", "ankle"))
    pca$mds_df[,"Final_Group"] <- factor(pca$mds_df[,"Final_Group"],
                                       levels=c("Healthy","VeRA","RA","OA"))

    # Build plot
    plt <- ggplot(pca$mds_df, aes(x=Dim1, y=Dim2, fill=final_location)) +
      geom_point(size=3, alpha=1, shape=21, stroke=0.3, color="black") +
      scale_fill_manual(values=colors) +
      xlab(paste0("PC1 (", pca$variance[1], "%)")) +
      ylab(paste0("PC2 (", pca$variance[2], "%)")) +
      theme_publish() +
      coord_fixed() +
      labs(fill="")
  
    # Save
    ggsave(paste0("chromhmm tracks/pca_", site_name, "_open_regions.png"),
           plot=plt, dpi=300, width=5, height=5, bg="transparent")
    ggsave(paste0("chromhmm tracks/pca_", site_name, "_open_regions.pdf"),
           plot=plt, dpi=300, width=5, height=5)

  }
  return(pca)
}
#-----------------------------------------------------------
# Run PCA analyses
#-----------------------------------------------------------
pca_all <- run_site_pca_plot("all_sites", site_data_all, patients, location_colors)
pca_knee <- run_site_pca_plot("knee", site_data_knee, patients, group_colors, site_label="knee")
pca_hand <- run_site_pca_plot("hand", site_data_hand, patients, group_colors, site_label="hand")
pca_shoulder <- run_site_pca_plot("shoulder", site_data_shoulder, patients, group_colors, site_label="shoulder")
```
