---
editor_options: 
  markdown: 
    wrap: 72
---

#DNA Methylation Differential Analysis R - analysis with integrated beta-values and Y chromosome probes filtered

```{r message=FALSE, warning=FALSE, results = 'hide'}
version
if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install()
```

##Load the libraries

```{r message=FALSE, warning=FALSE, results = 'hide'}
#loading those libraries
library(pheatmap)
library(gt)
library(ggvenn)
library(tidyverse)
library(viridis)
library(ggbreak)
library(minfi)
library(limma)
library(missMethyl)
library(minfiData)
library(DMRcate)
library(RColorBrewer)
library(dplyr)
library(readr)
library(ggrepel)
library(ggthemr)
library(EnhancedVolcano)
library(devtools)
library(GenomicRanges)
library(envalysis)
library(methylclock)
library(readxl)
library(VennDiagram)
library(IlluminaHumanMethylationEPICv2anno.20a1.hg38)
library(IlluminaHumanMethylation450kanno.ilmn12.hg19)
library(AnnotationDbi)
library(mCSEA)
library(rtracklayer)
select <- dplyr::select

```

#DMPs with chromHMM open regions - X & Y chromosome removed probes (all)
use the probes that overlap with the open regions across all samples - knee, shoulder, hand etc.
from chromHMM tracks
Then run DMP analysis across joints open region probes - 
sex and joint is a covariate 

```{r}
# -----------------------------
# Differential methylation analysis (ChromHMM-filtered probes)
# -----------------------------

setwd("/Users/egeezen/Desktop/PhD/2025 First Term/DNA Methylation/")

# -----------------------------
# Load data
# -----------------------------
final_data <- readRDS("merged_beta_values_and_annotation_file_after_filtered_xy.rds")
beta_values <- final_data$merged_beta_values
annotation_file <- final_data$annotation_file
patients <- final_data$patients

# Load ChromHMM-filtered probes (character vector)
probes_keep <- readRDS("chromhmm tracks/open_probes_all_sites_xy_removed.rds")

# -----------------------------
# Filter matrices
# -----------------------------
beta_values_filt <- beta_values[rownames(beta_values) %in% probes_keep, , drop = FALSE]
annotation_file_filt <- annotation_file[rownames(annotation_file) %in% probes_keep, , drop = FALSE]

# -----------------------------
# Convert beta to M-values
# -----------------------------
beta_to_m <- function(beta, offset = 1e-6) log2((beta + offset) / (1 - beta + offset))
m_values_filt <- beta_to_m(beta_values_filt)

# -----------------------------
# Save filtered data (beta_values)
# -----------------------------

filtered_data <- list(
  beta_values_filtered = beta_values_filt,
  annotation_file_filtered = annotation_file_filt,
  patients = patients
)

saveRDS(filtered_data, "outputs - after xy removal - open regions/open_region_filtered_data_all.rds")

# -----------------------------
# limma design
# -----------------------------
patients$Final_Group <- factor(patients$Final_Group, levels = c("Healthy", "VeRA", "RA", "OA"))
patients$final_location <- factor(patients$final_location)
patients$sex <- factor(patients$sex)

design <- model.matrix(
  ~ 0 + Final_Group + final_location + sex,
  data = patients
)

colnames(design) <- gsub("Final_Group", "", colnames(design))

# Fit model
fit <- lmFit(m_values_filt, design)

# Contrast matrix
contrast_matrix <- makeContrasts(
  RA_vs_Healthy = RA - Healthy,
  VeRA_vs_Healthy = VeRA - Healthy,
  OA_vs_Healthy = OA - Healthy,
  RA_vs_OA = RA - OA,
  levels = design
)

fit2 <- contrasts.fit(fit, contrast_matrix)
fit2 <- eBayes(fit2, trend = TRUE, robust = TRUE)

summary(decideTests(fit2))


# -----------------------------
# Extract DMPs
# -----------------------------
get_dmp_df <- function(coef_name, first_group, second_group) {
  DMPs_df <- topTable(fit2, coef = coef_name, number = Inf, sort.by = "P",
                      p.value = Inf, adjust.method = "BH")
  probes <- rownames(DMPs_df)

  group1 <- patients$Patient_Name[patients$Final_Group == first_group]
  group2 <- patients$Patient_Name[patients$Final_Group == second_group]

  deltaB <- rowMeans(beta_values_filt[probes, group1, drop = FALSE]) -
            rowMeans(beta_values_filt[probes, group2, drop = FALSE])

  DMPs_df$deltaB <- deltaB[probes]
  return(DMPs_df)
}

RA_vs_Healthy_DMPs <- get_dmp_df("RA_vs_Healthy", "RA", "Healthy")
RA_vs_OA_DMPs      <- get_dmp_df("RA_vs_OA", "RA", "OA")
OA_vs_Healthy_DMPs <- get_dmp_df("OA_vs_Healthy", "OA", "Healthy")
VeRA_vs_Healthy_DMPs <- get_dmp_df("VeRA_vs_Healthy", "VeRA", "Healthy")

# -----------------------------
# Filter significant DMPs (FDR)
# -----------------------------
filter_with_FDR <- function(DMPs) {
  DMPs %>%
    filter(adj.P.Val <= 0.05 ) %>%
    merge(annotation_file_filt, by = "row.names", all.x = TRUE)
}

RA_vs_Healthy_DMPs_FDR_filt <- filter_with_FDR(RA_vs_Healthy_DMPs)
OA_vs_Healthy_DMPs_FDR_filt <- filter_with_FDR(OA_vs_Healthy_DMPs)
VeRA_vs_Healthy_DMPs_FDR_filt <- filter_with_FDR(VeRA_vs_Healthy_DMPs)
RA_vs_OA_DMPs_FDR_filt <- filter_with_FDR(RA_vs_OA_DMPs)

count_dmps <- function(df, label) {
  data.frame(
    Contrast = label,
    Direction = c("Hyper", "Hypo"),
    Count = c(
      sum(df$adj.P.Val <= 0.05 & df$deltaB > 0),
      sum(df$adj.P.Val <= 0.05 & df$deltaB < 0)
    )
  )
}

plot_df <- bind_rows(
  count_dmps(RA_vs_Healthy_DMPs, "RA vs Healthy"),
  count_dmps(OA_vs_Healthy_DMPs, "OA vs Healthy"),
  count_dmps(VeRA_vs_Healthy_DMPs, "VeRA vs Healthy"),
  count_dmps(RA_vs_OA_DMPs, "RA vs OA")
)
  
plot_mirror <- plot_df %>%
  mutate(DisplayCount = ifelse(Direction == "Hypo", -Count, Count))
# 2. Plot
mirror_p <- ggplot(plot_mirror, aes(x = Contrast, y = DisplayCount, fill = Direction)) +
  geom_bar(stat = "identity", width = 0.7) +
  scale_fill_viridis(discrete = TRUE, option = "viridis", begin = 0.4, end = 0.7) +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.5) +
  scale_y_break(c(1000, 14500)) +
  scale_y_continuous(
    breaks = c(-500, 0, 500, 1000, 15000), 
    labels = abs
  ) +
  #scale_y_continuous(labels = abs) +
  labs(y = "FDR-significant CpGs", x = NULL) +
  theme_publish(base_size = 14, base_linewidth = 0.6) +
  theme(
    axis.text.x = element_text(angle = 35, hjust = 1),
    legend.position = "right"
  )

ggsave(filename = "fdr_mirrorplot.pdf",mirror_p, width = 5, height = 5)

# -----------------------------
# Filter significant DMPs (nominal pvalue < 0.01)
# -----------------------------

filter_for_exploratory <- function(DMPs) {
  DMPs %>%
    filter(P.Value <= 0.01 & abs(deltaB) >= 0.1) %>%
    merge(annotation_file_filt, by = "row.names", all.x = TRUE)
}

RA_vs_Healthy_DMPs_significant <- filter_for_exploratory(RA_vs_Healthy_DMPs)
RA_vs_OA_DMPs_significant      <- filter_for_exploratory(RA_vs_OA_DMPs)
OA_vs_Healthy_DMPs_significant <- filter_for_exploratory(OA_vs_Healthy_DMPs)

# -----------------------------
#do a venn plot

list_for_venn <- list(
  "RA vs Healthy" = RA_vs_Healthy_DMPs_significant$Row.names,
  "OA vs Healthy" = OA_vs_Healthy_DMPs_significant$Row.names)

venn_p <- ggvenn(
  list_for_venn,
  fill_color = c("#A888B5", "#C5D3E8"),
  fill_alpha = 0.6,
  stroke_color = "white",
  stroke_size = 0.8,
  set_name_size = 5,
  text_size = 4.5
) +
  labs(title = "Overlap of Exploratory DMPs (p < 0.01)") +
  theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))

ggsave(filename = "venn_plot_ra_oa_dmps.pdf",venn_p, width = 5, height = 5)

# -----------------------------
#do a barplot

# Hyper DMPs
hyper_ra <- RA_vs_Healthy_DMPs_significant %>% filter(deltaB > 0) %>% pull(Row.names)
hyper_oa <- OA_vs_Healthy_DMPs_significant %>% filter(deltaB > 0) %>% pull(Row.names)

# Hypo DMPs
hypo_ra <- RA_vs_Healthy_DMPs_significant %>% filter(deltaB < 0) %>% pull(Row.names)
hypo_oa <- OA_vs_Healthy_DMPs_significant %>% filter(deltaB < 0) %>% pull(Row.names)

# Compute shared and unique counts
hyper_shared <- length(intersect(hyper_ra, hyper_oa))
hyper_ra_only <- length(setdiff(hyper_ra, hyper_oa))
hyper_oa_only <- length(setdiff(hyper_oa, hyper_ra))

hypo_shared <- length(intersect(hypo_ra, hypo_oa))
hypo_ra_only <- length(setdiff(hypo_ra, hypo_oa))
hypo_oa_only <- length(setdiff(hypo_oa, hypo_ra))

# Create a dataframe for plotting
plot_df <- data.frame(
  Methylation = rep(c("Hyper", "Hypo"), each = 3),
  Categories = rep(c("RA only", "OA only", "Shared"), 2),
  Count = c(hyper_ra_only, hyper_oa_only, hyper_shared,
            hypo_ra_only, hypo_oa_only, hypo_shared)
) %>%
  group_by(Methylation) %>%
  mutate(
    Total = sum(Count),
    Percent = Count / Total * 100,
    Label = sprintf("%d\n(%.1f%%)", Count, Percent)
  ) %>%
  ungroup()

# Plot
bar_plot <- ggplot(plot_df, aes(x = Methylation, y = Count, fill = Categories)) +
  geom_bar(stat = "identity") + 
  geom_text(
    aes(label = Label),
    position = position_stack(vjust = 0.5),
    size = 5
  ) +
  scale_fill_manual(values = c(
    "RA only" = "#A888B5",
    "OA only" = "#C5D3E8",
    "Shared" = "#7FBF7F"
  )) +
  ylab("Number of DMPs") +
  xlab("") +
  theme_publish(base_size = 15, base_linewidth = 0.6) +
  theme(axis.text.x = element_text(face = "bold"),
        legend.position = "right")

ggsave(filename = "bar_plot_ra_oa_dmps.pdf",bar_plot, width = 6, height = 6)

# -----------------------------
# Save DMP results
# -----------------------------
dmp_results <- list(
  "RA_vs_Healthy_DMPs" = RA_vs_Healthy_DMPs,
  "RA_vs_Healthy_DMPs_significant" = RA_vs_Healthy_DMPs_significant,
  "OA_vs_Healthy_DMPs" = OA_vs_Healthy_DMPs,
  "OA_vs_Healthy_DMPs_significant" = OA_vs_Healthy_DMPs_significant,
  "RA_vs_OA_DMPs" = RA_vs_OA_DMPs,
  "RA_vs_OA_DMPs_significant" = RA_vs_OA_DMPs_significant
)
saveRDS(dmp_results, file = "outputs - after xy removal - open regions/RA_and_OA_vs_Healthy_filteredChromHMM.rds")

# -----------------------------
# Run DMRcate on filtered probes
# -----------------------------
contrast_names <- c("RA_vs_Healthy", "OA_vs_Healthy", "RA_vs_OA")
dmr_results <- list()

for (contrast in contrast_names) {
  message("Running DMRcate for: ", contrast)
  
  annotation <- cpg.annotate(
    object = as.matrix(m_values_filt),
    datatype = "array",
    what = "M",
    analysis.type = "differential",
    contrasts = TRUE,
    cont.matrix = contrast_matrix,
    design = design,
    coef = contrast,
    arraytype = "450K" 
  )
  
  dmrc_out <- dmrcate(annotation, lambda = 1000, C = 2, pcutoff = 0.05)
  dmr_results[[contrast]] <- extractRanges(dmrc_out, genome = "hg19")
}

# -----------------------------
# LiftOver to hg38
# -----------------------------
chain <- import.chain("~/Desktop/PhD/2024 First Term/DNA Methylation/Merged_samples/hg19ToHg38.over.chain")

dmr_hg38 <- lapply(dmr_results, function(gr) {
  unlist(liftOver(gr, chain))
})

saveRDS(dmr_hg38, "outputs - after xy removal - open regions/RA_vs_Healthy_and_OA_vs_Healthy_dmr_results_filteredChromHMM.rds")

```

##Save BED files for IGV

```{r}
# --------------------------------------------------
# Helper function to export DMPs to BED for IGV
# --------------------------------------------------
# Assumptions:
# - your DMP table contains chromosome and genomic position columns
# - common column names handled below
# - rownames are CpG IDs if no explicit probe column exists

#read the dmps
dmp_results <- readRDS(file = "outputs - after xy removal - open regions/RA_and_OA_vs_Healthy_filteredChromHMM.rds")

export_dmps_to_bed <- function(dmp_df, out_file, genome_build = "hg38") {
  
  df <- dmp_df %>%
    as.data.frame() %>%
    rownames_to_column(var = "CpG")
  
  # Try to detect chromosome column
  chr_col <- c("chr", "CHR", "Chromosome", "chromosome", "seqnames")[
    c("chr", "CHR", "Chromosome", "chromosome", "seqnames") %in% colnames(df)
  ][1]
  
  # Try to detect position column
  pos_col <- c("pos", "POS", "Position", "position", "start", "Start", "MAPINFO")[
    c("pos", "POS", "Position", "position", "start", "Start", "MAPINFO") %in% colnames(df)
  ][1]
  
  if (is.na(chr_col) || is.na(pos_col)) {
    stop("Could not detect chromosome and/or position columns. Please check your DMP table column names.")
  }
  
  # Pick useful metadata columns if present
  extra_cols <- c("logFC", "delta_beta", "Delta_Beta", "P.Value", "adj.P.Val", "gene")
  extra_cols <- extra_cols[extra_cols %in% colnames(df)]
  
  bed_df <- df %>%
    mutate(
      chr = .data[[chr_col]],
      pos = as.integer(.data[[pos_col]])
    ) %>%
    filter(!is.na(chr), !is.na(pos)) %>%
    mutate(
      chr = ifelse(grepl("^chr", chr), chr, paste0("chr", chr)),
      start = pos - 1,
      end = pos
    ) %>%
    select(chr, start, end, CpG, all_of(extra_cols)) %>%
    arrange(chr, start)
  
  write_tsv(bed_df, out_file, col_names = FALSE)
  
  message("Saved BED file: ", out_file)
  message("Genome build assumed: ", genome_build)
}

# --------------------------------------------------
# Export only significant RA vs Healthy and OA vs Healthy DMPs
# --------------------------------------------------

export_dmps_to_bed(
  RA_vs_Healthy_DMPs_significant,
  out_file = "outputs - after xy removal - open regions/RA_vs_Healthy_significant_DMPs_IGV.bed"
)

export_dmps_to_bed(
  OA_vs_Healthy_DMPs_significant,
  out_file = "outputs - after xy removal - open regions/OA_vs_Healthy_significant_DMPs_IGV.bed"
)
```

#DMPs with chromHMM open regions - joint comparisons
Only compare joints to obtain joint specific DMRs (within OA + RA comparing upper extremity to lower extremity)

```{r}
# ============================================================
# RA + OA only: Upper vs Lower extremity (ChromHMM-open probes)
# ============================================================

setwd("/Users/egeezen/Desktop/PhD/2025 First Term/DNA Methylation/")

# -----------------------------
# Load data
# -----------------------------
final_data <- readRDS("merged_beta_values_and_annotation_file_after_filtered_xy.rds")
beta_values <- final_data$merged_beta_values
annotation_file <- final_data$annotation_file
patients <- final_data$patients

probes_keep <- readRDS("chromhmm tracks/open_probes_all_sites_xy_removed.rds")

# -----------------------------
# Filter to ChromHMM-open probes
# -----------------------------
beta_values_filt <- beta_values[rownames(beta_values) %in% probes_keep, , drop = FALSE]
annotation_file_filt <- annotation_file[rownames(annotation_file) %in% probes_keep, , drop = FALSE]

beta_to_m <- function(beta, offset = 1e-6) log2((beta + offset) / (1 - beta + offset))
m_values_filt <- beta_to_m(beta_values_filt)

# ============================================================
# 1) Subset to RA + OA only, define Upper/Lower extremity
# ============================================================

# Keep only RA and OA
patients2 <- patients %>%
  filter(Final_Group %in% c("RA", "OA"))

# Define upper vs lower from your location variable
upper_levels <- c("hand", "shoulder")
lower_levels <- c("knee") 

patients2 <- patients2 %>%
  mutate(
    final_location = as.character(final_location),
    extremity = case_when(
      final_location %in% upper_levels ~ "Upper",
      final_location %in% lower_levels ~ "Lower",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(extremity)) %>%
  mutate(
    Final_Group = factor(Final_Group, levels = c("OA","RA")),  # reference OA first (arbitrary)
    extremity   = factor(extremity, levels = c("Lower","Upper")),
    sex         = factor(sex)
  )

# Make sure columns match samples
# Here you used Patient_Name to index the columns. Keep that convention:
stopifnot(all(patients2$Patient_Name %in% colnames(m_values_filt)))

m_values2 <- m_values_filt[, patients2$Patient_Name, drop = FALSE]
beta_values2 <- beta_values_filt[, patients2$Patient_Name, drop = FALSE]

# ============================================================
# 2) limma: Upper vs Lower, adjusting for Diagnosis + Sex
#    This asks: "Upper-Lower within RA+OA, controlling for RA/OA"
# ============================================================

design_UL <- model.matrix(~ 0 + extremity + Final_Group + sex, data = patients2)
colnames(design_UL) <- gsub("^extremity", "", colnames(design_UL))

fit_UL <- lmFit(m_values2, design_UL)

contrast_UL <- makeContrasts(
  Upper_vs_Lower = Upper - Lower,
  levels = design_UL
)

fit2_UL <- contrasts.fit(fit_UL, contrast_UL)
fit2_UL <- eBayes(fit2_UL, trend = TRUE, robust = TRUE)

summary(decideTests(fit2_UL))

# ============================================================
# 3) Extract DMPs for Upper_vs_Lower (with deltaB)
# ============================================================

get_dmp_df_UL <- function(coef_name = "Upper_vs_Lower") {
  dmp <- topTable(fit2_UL, coef = coef_name, number = Inf, sort.by = "P",
                  p.value = Inf, adjust.method = "BH")
  probes <- rownames(dmp)

  upper_samps <- patients2$Patient_Name[patients2$extremity == "Upper"]
  lower_samps <- patients2$Patient_Name[patients2$extremity == "Lower"]

  deltaB <- rowMeans(beta_values2[probes, upper_samps, drop = FALSE]) -
            rowMeans(beta_values2[probes, lower_samps, drop = FALSE])

  dmp$deltaB <- deltaB[probes]
  dmp
}

UL_DMPs <- get_dmp_df_UL("Upper_vs_Lower")

# Exploratory set (same as the diagnosis comparisons: p<0.01 & |deltaB|>=0.1)
UL_DMPs_expl <- UL_DMPs %>%
  filter(P.Value <= 0.01 & abs(deltaB) >= 0.1) %>%
  merge(annotation_file_filt, by = "row.names", all.x = TRUE)

# Save
saveRDS(
  list(
    patients_RA_OA = patients2,
    UL_DMPs = UL_DMPs,
    UL_DMPs_expl = UL_DMPs_expl
  ),
  file = "outputs - after xy removal - open regions/RA_OA_upper_vs_lower_DMPs_filteredChromHMM.rds"
)

# ============================================================
# 4) Run DMRcate for Upper_vs_Lower on the same filtered probes
# ============================================================

annotation_UL <- cpg.annotate(
  object = as.matrix(m_values2),
  datatype = "array",
  what = "M",
  analysis.type = "differential",
  design = design_UL,
  contrasts = TRUE,
  cont.matrix = contrast_UL,
  coef = "Upper_vs_Lower",
  arraytype = "450K"
)

dmrc_UL <- dmrcate(annotation_UL, lambda = 1000, C = 2, pcutoff = 0.05)
dmr_UL_hg19 <- extractRanges(dmrc_UL, genome = "hg19")

# LiftOver to hg38 (same as you)
chain <- import.chain("~/Desktop/PhD/2024 First Term/DNA Methylation/Merged_samples/hg19ToHg38.over.chain")
dmr_UL_hg38 <- unlist(liftOver(dmr_UL_hg19, chain))

saveRDS(
  list(
    Upper_vs_Lower_hg19 = dmr_UL_hg19,
    Upper_vs_Lower_hg38 = dmr_UL_hg38
  ),
  "outputs - after xy removal - open regions/RA_OA_upper_vs_lower_DMRs_filteredChromHMM.rds"
)

```


